---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](LICENSE.md)

## MPB grcMalaria Package
This package, inspired by the malariagen/grcMalaria project, designed for analyzing SpotMalaria Genetic Report Cards (GRC). The package provides a comprehensive toolkit for population genetics analysis, seamlessly integrating spatial data and visualization to facilitate the following:

* **Prevalence and Diversity Analysis:** Gain insights into the prevalence and genetic diversity of malaria parasites across various locations.
* **Drug Resistance Profiling:** Detect and characterize drug resistance patterns in malaria strains.
* **Geographical Mapping:** Visualize the distribution and movement of malaria strains across different locations..
* **Genetic Relationship Exploration:** Examine genetic relatedness among samples using IBS (Identity-By-State) matrices and associated plots.

Below is a simple workflow example demonstrating the main functions in the MPB grcMalaria package.

### Creating the Final GRC Data

The combine_grc_sheets function consolidates multiple SpotMalaria Genetic Report Cards (GRC) into a single data set for downstream analysis. This function also allows you to filter the data by country.

```{r eval = FALSE}
grc_data <-
  combine_grc_sheets(
    input_folder = "C:/Users/bngwa/Documents/Brandon/GDA_Markdown/All_GRC_Reads_Gambia",
    country = "Gambia",
    save_output = TRUE,
    output_dir = "C:/Users/bngwa/Videos"
  )
```


### Create the drug condition columns

Use the gene_classifier function to classify samples into categories such as **Resistant**, **Sensitive**, **Mixed_Resistance**, **Undetermined** and **Missing** based on the haplotypes of certain gene markers associated with specific drugs, e.g, PfCRT which is associated with Chloroquine. New columns are then created using the name of each drug.

```{r eval = FALSE}
grc_data <-
  gene_classifier(
    df = grc_data,
    drug_column = "Chloroquine",
    save_output = TRUE
  )
```


### Preparing Metadata for Mapping

Integrate shapefile data and geographic coordinates to create a list using mapping_data. This is necessary for creating spatial visualizations subsequent analyses.

```{r, echo = TRUE, eval = FALSE}
# import your geographical data for mapping
gmb_shpfile <- st_read(system.file("extdata", "geoBoundaries-GMB-ADM3_simplified.shp", package = "grcMPB"))
longitude_latitude <- read_excel(system.file("extdata", "LongLat_data.xlsx", package = "grcMPB"))

geo_data <-
  mapping_data(
    shapefile = gmb_shpfile,
    long_lat_data = longitude_latitude,
    location_col = "Location",
    long_col = "long",
    lat_col = "lat"
  )
```

### Setting Time Periods

Define specific time periods to filter data when generating plots. Below is an example of a list specifying individual years and ranges.

```{r}
## You can create a single or range of years to filter buy before generating the plots
periods <-
  list(
    list(name = "2021", type = "year", start = "2021"),
    list(name = "2017-19", type = "period", start = "2017", end = "2019")
  )
```


### Generating a Sample Count Map

Create a map displaying sample counts across different locations.

```{r, echo = TRUE, eval = FALSE}
sample_count_map(
  df = grc_data,
  map_data = geo_data,
  time = NULL,
  breaks = c(10, 100, 200, 300),
  label_size = 2.5,
  scale_circle_size = 11,
  save_output = TRUE
)
```

